Synthesizing Quotient Lenses

### 2018-03-06 papers/QuotientLenses/main.tex
merged Benjamins changes into mine

One way\bcp{Do you mean to highlight the suggestion that there are other
  ways too?} to classify lenses is by the laws they obey.  Amongst the most
restrictive lenses are the \emph{bijective lenses}.  These lenses demand
that get and put functions form a bijection, meaning no information
is lost when one translates from one data format to another.  While this
strong guarantee can be useful in some contexts, it is more often the case
that some of the information in one or both of the data formats is superfluous.
For instance, one typically does not care to preserve the amount of
whitespace between characters.  In other circumstances, capitalization
or the order of fields in a record or items in a list may be unimportant.
%
%bijective lenses have
%strong correctness guarantees, they are often
%However, these lens laws are too restrictive for many important transformations.
%The bijective lens laws do not permit transformations that ignore this
%unimportant information---.
Many data formats have other ad hoc rules as well.
In bibliographies, for example, recall that in \bibtex{},
an author can be specified as
``\cd{lastname, firstname}'' or as ``\cd{firstname lastname}''---it does not matter which of the
two formats is used.
%While it is important to allow authors
%to be input in both ways, we would like these authors to only be output as
%\cd{firstname lastname}, but the lens laws prevent such functions.
%
{\em Quotient lenses}~\cite{quotientlenses} loosen many of the restrictions
of bijective lenses by introducing an equivalence relation\bcp{actually,
  two: one on each side; but more importantly ``introducing an equivalence
  relation'' is rather vague.} and considering
bijections modulo that relation.  Such quotient lenses make it possible
to align many more data sources than would otherwise be possible by
ignoring details that a programmer views as inessential.

Ordinary bijective lenses are fairly tricky to program\bcp{do we have
  concrete evidence for this?}, especially for
beginners\bcp{or this?}, as they require programmers to think in two
directions at once\bcp{my own experience suggests that this is the less
  important issue...}
and learn a new type system, which comes with some tricky ambiguity
constraints\bcp{...and this is the bigger one.  But we don't explain what
  these ambiguity constraints are or why they must be imposed.}.
Unfortunately, quotient lenses do not improve the situation\bcp{Again, it
  feels rather negative.}:  To
write a quotient lens in Boomerang, programmers must work with standard lens
combinators as well as a system of \emph{canonizers}.
These canonizers encode two functions, one that
maps data represented in the original external format
to a canonical element, and another, in the opposite direction, that chooses
which elements of the original format correspond to values in the canonical
format. Furthermore, if a user wants to define the domain and range of
a lens explicitly (a programming practice that is encouraged), she must
provide regular expressions defining input and output formats.
Consequently, to define a quotient lens, one must often
construct three different sorts
of terms: (1) regular expressions that define the two data formats
to be processed, (2) canonizers to normalize the data
and choose representatives, and (3) a bijective lens that maps
one canonical form into another.  Somewhat frustratingly,
these three terms share a fair amount of structure.
%Moreover, writing the lens between the equivalence classes can be
%challenging due to the rather fiddly typechecking, mainly due to
%``non-ambiguity'' conditions in the lens well-formedness rules.

@@

\section{Conclusion}
\bcp{And future work!}  \saz{I'm lukewarm about including a description of
  future work.  Unless there are specific conjectures or open problems that
  are worth mentioning, I'm not sure that it's really that important.}
\bcp{Agreed. But don't we have any ideas for how to extend or generalize or
  things to do next??  E.g., is it completely trivial to extend all this to
  asymmetric and symmetric lenses?  (If so, we should say so!)  Do we have
  more ideas for QRE primitives?  Etc.}
\label{concl}

>>>>

One way\bcp{Do you mean to highlight the suggestion that there are other
  ways too?} to classify lenses is by the laws they obey.  Amongst the most
restrictive lenses are the \emph{bijective lenses}.  These lenses demand
that get and put functions form a bijection, meaning no information
is lost when one translates from one data format to another.  While this
strong guarantee can be useful in some contexts, it is more often the case
that some of the information in one or both of the data formats is superfluous.
For instance, one typically does not care to preserve the amount of
whitespace between characters.  In other circumstances, capitalization
or the order of fields in a record or items in a list may be unimportant.
%
%bijective lenses have
%strong correctness guarantees, they are often
%However, these lens laws are too restrictive for many important transformations.
%The bijective lens laws do not permit transformations that ignore this
%unimportant information---.
Many data formats have other ad hoc rules as well.
In bibliographies, for example, recall that in \bibtex{},
an author can be specified as
``\cd{lastname, firstname}'' or as ``\cd{firstname lastname}''---it does not matter which of the
two formats is used.
%While it is important to allow authors
%to be input in both ways, we would like these authors to only be output as
%\cd{firstname lastname}, but the lens laws prevent such functions.
%
{\em Quotient lenses}~\cite{quotientlenses} loosen many of the restrictions
of bijective lenses by introducing an equivalence relation\bcp{actually,
  two: one on each side; but more importantly ``introducing an equivalence
  relation'' is rather vague.} and considering
bijections modulo that relation.  Such quotient lenses make it possible
to align many more data sources than would otherwise be possible by
ignoring details that a programmer views as inessential.
Ordinary bijective lenses are fairly tricky to program\bcp{do we have
  concrete evidence for this?}, especially for
beginners\bcp{or this?}, as they require programmers to think in two
directions at once\bcp{my own experience suggests that this is the less
  important issue...}
and learn a new type system, which comes with some tricky ambiguity
constraints\bcp{...and this is the bigger one.  But we don't explain what
  these ambiguity constraints are or why they must be imposed.}.
Unfortunately, quotient lenses do not improve the situation\bcp{Again, it
  feels rather negative.}:  To
write a quotient lens in Boomerang, programmers must work with standard lens
combinators as well as a system of \emph{canonizers}.
These canonizers encode two functions, one that
maps data represented in the original external format
to a canonical element, and another, in the opposite direction, that chooses
which elements of the original format correspond to values in the canonical
format. Furthermore, if a user wants to define the domain and range of
a lens explicitly (a programming practice that is encouraged), she must
provide regular expressions defining input and output formats.
Consequently, to define a quotient lens, one must often
construct three different sorts
of terms: (1) regular expressions that define the two data formats
to be processed, (2) canonizers to normalize the data
and choose representatives, and (3) a bijective lens that maps
one canonical form into another.  Somewhat frustratingly,
these three terms share a fair amount of structure.
%Moreover, writing the lens between the equivalence classes can be
%challenging due to the rather fiddly typechecking, mainly due to
%``non-ambiguity'' conditions in the lens well-formedness rules.

@@

\section{Conclusion}\bcp{And future work!!!}  \saz{I'm luke warm about including
  a description of future work.  Unless there are specific conjectures or open
  problems that are worth mentioning, I'm not sure that it's really that important.}
\label{concl}
